The Diminishing Returns of Music in the Digital Era

### 2024-04-18 paper/paper.qmd
final edits :)

library(tidyverse)
library(knitr)
library(readr)
library(ggplot2)
library(brms)
```

@@

This paper seeks to explain the economic narrative of the music industry's journey through the digital age. By analyzing a dataset on music sales, we unravel how the industry's revenue per unit has been affected since the decline of physical media. The estimand in this paper is the average annual revenue generated by each music format over time. Amidst the backdrop of booming digital consumption, we find an inverse trajectory in revenue, raising questions about the long-term sustainability of current digital revenue models.

@@

The structure of the paper is as follows: we start with a detailed examination of the dataset, charting the industry's sales and revenue trends (@sec-data). Subsequent sections present our analysis of the data (@sec-results), revealing significant insights into the economic shifts within the industry. We then pivot to a discussion (@sec-discussion), contextualizing our findings within the broader framework of contemporary media consumption. The analysis was conducted using the statistical programming language R (@citeR), utilizing packages such as rstanarm (@rstanarm), tidyverse (@tidyverse), knitr (@knitr), readr (@readr), ggplot2 (@gg), brms (@brms), and additional papers such as "The Recording Industry's Digital Dilemma: Challenges and Opportunities in High Piracy Markets"(@Keintz2008TheRI), "When Business Models Go Bad: The Music Industry's Future" (@Wilde2004WhenBM), "Realizing the social and commercial potential of interactive technologies" (@Mundorf2002RealizingTS), and "Study on the New Models of Music Industry in the Era of AI and Blockchain" (@Shang2020StudyOT).

@@

## Source {#sec-source}
The main data source is a dataset from the Recording Industry Association of America (RIAA) (@RIAA), tracking recorded music revenues by format in the United States from 1973 to 2019. This dataset was generated by Matt Bass and has been adjusted for inflation, offering a standardized monetary comparison across five decades of music industry economics. Since the RIAA has yet to properly respond to requests for the use of their dataset, we made use of a Kaggle dataset from Kaggle user "Larxel" that directly cites the RIAA dataset and uses variables and other data from it. There could have been other datasets from Kaggle used but they weren't because they didn't cite the RIAA dataset.

Detailing yearly revenue across a spectrum of music formats, the RIAA dataset offers a look at the industry's economic fluctuations, tracing the lineage from traditional mediums like LPs and cassettes to modern digital forms such as streams and downloads. This level of detail affords an understanding of changing consumer preferences and technological advancements that have influenced the music market in the United States.

The RIAA dataset was specifically chosen for its focus on the U.S. market and its longitudinal data, allowing for an in-depth analysis of market shifts over fifty years. Other global datasets did not offer the same level of historical depth or the focused granularity required to assess the U.S. market's unique journey through digital transformation. The RIAA dataset is composed of multiple variables that encapsulate different facets of the music industry. Variables such as "Format", "Year", and "Value" have been carefully vetted to ensure consistent and accurate representations of the industry's financial outlook over time. This extensive examination guarantees that the paper rests on a solid empirical foundation, capable of yielding substantial insight into the evolution of the music industry.

@@

@@

music_data <- read_csv("/Users/victortimir/music-industry-analysis/data/raw_data/musicdata-1.csv")

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-averagerevenue
#| tbl-cap: "Music Industry Average Revenue per Format"
#| warning: false
#| results: asis
#| 
music_data <- read_csv("/Users/victortimir/music-industry-analysis/data/raw_data/musicdata-1.csv")

@@

@@

The goal of our modelling strategy is twofold. Firstly, we seek to evaluate how different music formats contribute to overall industry revenue over time. This involves analyzing the changes in revenue streams from physical sales such as CDs and vinyl to digital formats like streaming and downloads. Secondly, we seek to use these insights to forecast future revenue trends for these formats.

@@

music_data <- read_csv("/Users/victortimir/music-industry-analysis/data/raw_data/musicdata-1.csv")

bayesian_model <- brm(value_actual ~ format, data = music_data, 
                      family = gaussian(), 
                      prior = c(set_prior("normal(0, 10)", class = "b"),
                                set_prior("cauchy(0, 2)", class = "sigma")),
                      chains = 4, iter = 2000, warmup = 500, 
                      control = list(adapt_delta = 0.95),
                      refresh = 0)
```

@@

What I believe to be the solution to the music industry's diminishing returns lies with both sentiments. To succeed, there must be a new, convenient business model that offers a compelling value for digital music consumption and an extensive library of music to boot. From the artist's perspective, touring has become prohibitively expensive and revenue models tied to streaming services fail to properly compensate them for their work. Due to social media engagement and streaming services, artists are also incentivized to release more music at a faster rate to satisfy each respective algorithm and model. Drake, one of the biggest artists of the 21st century, has arguably benefited the most from a mastery of social media algorithms and streaming service models, as he took advantage of his pre-streaming era success to release lengthy projects at an annual rate to maximize streams and increase his social media engagement. In 2013, Thom Yorke from Radiohead pulled all of his own (not counting Radiohead) songs from Spotify in protest at how much it pays artists. In 2007, Radiohead released their album "In Rainbows" independently as a "pay-what-you-want" download, meaning that fans could set their price. The band's reasoning for this was that it "liberated them from conventional promotional formats and removed barriers to audiences". Services like Bandcamp offer the same thing for smaller artists. If this business model was adopted by larger artists and properly integrated with smartphones and computers, it would offer enough convenience to the consumer to be a reliable alternative that would also generate revenue for artists and the industry as a whole, as it challenges old models and finds new ways to connect with fans.

@@

The effect of Tiktok on the music industry can't have been explicitly applied to the present data in this paper because this data cuts off in 2019 and Tiktok started to gain prominence in 2018. However, I think Tiktok perfectly illustrates the quality of the user experience and its subsequent effect on the music industry. The user-generated content that TikTok is made up of has opened new doors for revenue through advertising partnerships and virality in addition to some content monetization. Mainly, the platform serves as a marketing tool, enabling artists to build their brand and engage with fans directly. Collectively, these digital alternatives complement streaming by diversifying the revenue portfolio of the industry, showcasing the multifaceted nature of music consumption in the digital age. However, the issue with this is that the complementation of streaming with TikTok has led to additional convenience for the consumer. This duo of streaming and TikTok leads to people consuming music as if it were free, as described by musician James Blake. He specifically argued that Tiktok has negatively affected songwriting by encouraging artists to write differently to appeal to the Tiktok algorithm and the attention deficit of the consumer by presenting a vast library of quick user-generated content tailor-made for the consumer.

@@

The analysis within this paper, while extensive, is not without limitations that should be acknowledged. Firstly, the dataset's restriction to revenue figures may overlook the varied consumer behaviors and preferences that drive the market. Future research could benefit from integrating qualitative data, such as consumer surveys, to complement the quantitative revenue analysis. Additionally, the rapid pace of technological change means that the dataset may quickly become outdated; continuous data collection and analysis will be important to maintain relevance. The dataset only includes information on North America. Incorporating worldwide data with @Keintz2008TheRI suggestions could lead to additional substantial observations.

Moreover, the current dataset does not capture the entire scope of the music industry and the revenue it generates from other streams of revenue such as live performances, merchandise sales, and licensing deals. Future iterations of this research should aim to include these aspects for a more complete view of the industry's economic outlook.

The industry is also moving towards more immersive and interactive experiences, such as virtual concerts or Twitch streams, areas not covered by the present dataset. Including these emerging trends could provide valuable insights into future revenue potentials and shifts in consumer engagement.

>>>>

library(tidyverse)
library(knitr)
library(readr)
library(ggplot2)
library(brms)
library(arrow)
```

@@

This paper seeks to explain the economic narrative of the music industry's journey through the digital age. By analyzing a dataset on music sales, we unravel how the industry's revenue per unit has been affected since the decline of physical media. The estimand in this paper is the average annual revenue generated by each music format over time. Amidst the backdrop of booming digital consumption, we find an inverse trajectory in revenue, raising questions about the long-term sustainability of current digital revenue models. Other papers have similar findings but they are much older and they attribute this trajectory to an increase in online piracy.

@@

The structure of the paper is as follows: we start with a detailed examination of the dataset, charting the industry's sales and revenue trends (@sec-data). Subsequent sections present our analysis of the data (@sec-results), revealing significant insights into the economic shifts within the industry. We then pivot to a discussion (@sec-discussion), contextualizing our findings within the broader framework of contemporary media consumption. The analysis was conducted using the statistical programming language R (@citeR), utilizing packages such as rstanarm (@rstanarm), tidyverse (@tidyverse), knitr (@knitr), readr (@readr), ggplot2 (@gg), brms (@brms), arrow (@arrow), sf (@sf), janitor (@janitor), testthat (@test) and additional papers such as "The Recording Industry's Digital Dilemma: Challenges and Opportunities in High Piracy Markets"(@Keintz2008TheRI), "When Business Models Go Bad: The Music Industry's Future" (@Wilde2004WhenBM), "Realizing the social and commercial potential of interactive technologies" (@Mundorf2002RealizingTS), and "Study on the New Models of Music Industry in the Era of AI and Blockchain" (@Shang2020StudyOT).

@@

## Source {#sec-source}
The main data source is a dataset from the Recording Industry Association of America (RIAA) (@RIAA), tracking recorded music revenues by format in the United States from 1973 to 2019. This dataset was generated by Matt Bass and has been adjusted for inflation, offering a standardized monetary comparison across five decades of music industry sales. Since the RIAA has yet to properly respond to requests for the use of their dataset, we made use of a Kaggle dataset from Kaggle user "Larxel" that directly cites the RIAA dataset and uses variables and other data from it. There could have been other datasets from Kaggle used but they weren't because they didn't cite the RIAA dataset.

Detailing yearly revenue across a spectrum of music formats, the RIAA dataset offers a look at the industry's economic fluctuations, tracing the lineage from traditional mediums like LPs and cassettes to modern digital forms such as streams and downloads. This level of detail affords an understanding of changing consumer preferences and technological advancements that have influenced the music market in the U.S.

The RIAA dataset was specifically chosen for its focus on the U.S. market and its longitudinal data, allowing for an in-depth analysis of market shifts over fifty years. Other global datasets did not offer the same level of historical depth or the focused granularity required to assess the U.S. market's unique journey through digital transformation. The RIAA dataset is composed of multiple variables that encapsulate different facets of the music industry. Variables such as "Format", "Year", and "Value" have been carefully vetted to ensure consistent and accurate representations of the industry's financial outlook over time. This examination guarantees that the paper rests on a solid empirical foundation, capable of yielding significant insight into the evolution of the music industry.

@@

The RIAA dataset is the definitive source of revenue data for the recorded music industry in the United States. All U.S. music industry sales data is directly reported to the RIAA. Then, it is compiled, cleaned, analyzed, and depending on certain certifications or awards, is finally processed. Their data is updated quarterly but in the case of this dataset, is updated annually. The data is immensely accurate, updated consistently, and handled in a way that respects privacy and confidentiality.

@@

music_data <- read_parquet("/Users/victortimir/music-industry-analysis/data/analysis_data/musicdata.parquet")

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-averagerevenue
#| tbl-cap: "Music Industry Average Revenue per Format"
#| warning: false
#| results: asis
#| 
music_data <- read_parquet("/Users/victortimir/music-industry-analysis/data/analysis_data/musicdata.parquet")

@@

Packages used to simulate, download, clean, and test the data include tidyverse (@tidyverse), ggplot2 (@gg), arrow (@arrow), sf (@sf), janitor (@janitor), and testthat (@test).

@@

The goal of our modeling strategy is twofold. Firstly, we seek to evaluate how different music formats contribute to overall industry revenue over time. This involves analyzing the changes in revenue streams from physical sales such as CDs and vinyl to digital formats like streaming and downloads. Secondly, we seek to use these insights to forecast future revenue trends for these formats.

@@

music_data <- read_parquet("/Users/victortimir/music-industry-analysis/data/analysis_data/musicdata.parquet")

bayesian_model <- brm(value_actual ~ format, data = music_data, 
                      family = gaussian(), 
                      prior = c(set_prior("normal(0, 10)", class = "b"),
                                set_prior("cauchy(0, 2)", class = "sigma")),
                      chains = 4, iter = 2000, warmup = 500, 
                      control = list(adapt_delta = 0.95),
                      refresh = 0)
saveRDS(bayesian_model, file = "second_model.rds")
```

@@

What I believe to be the solution to the music industry's diminishing returns lies with both sentiments. To succeed, there must be a new, convenient business model that offers a compelling value for digital music consumption and a sizeable library of music to boot. From the artist's perspective, touring has become prohibitively expensive and revenue models tied to streaming services fail to properly compensate them for their work. Due to social media engagement and streaming services, artists are also incentivized to release more music at a faster rate to satisfy each respective algorithm and model. Drake, one of the biggest artists of the 21st century, has arguably benefited the most from a mastery of social media algorithms and streaming service models, as he took advantage of his pre-streaming era success to release lengthy projects at an annual rate to maximize streams and increase his social media engagement. In 2013, Thom Yorke from Radiohead pulled all of his own (not counting Radiohead) songs from Spotify in protest at how much it pays artists. In 2007, Radiohead released their album "In Rainbows" independently as a "pay-what-you-want" download, meaning that fans could set their price. The band's reasoning for this was that it "liberated them from conventional promotional formats and removed barriers to audiences". Services like Bandcamp offer the same thing for smaller artists. If this business model was adopted by larger artists and properly integrated with smartphones and computers, it would offer enough convenience to the consumer to be a reliable alternative that would also generate revenue for artists and the industry as a whole, as it challenges old models and finds new ways to connect with fans.

@@

The effect of Tiktok on the music industry can't have been explicitly applied to the present data in this paper because this data cuts off in 2019 and Tiktok started to gain prominence in 2018. However, I think TikTok perfectly illustrates the quality of the user experience and its subsequent effect on the music industry. The user-generated content that TikTok is made up of has opened new doors for revenue through advertising partnerships and virality in addition to some content monetization. Mainly, the platform serves as a marketing tool, enabling artists to build their brand and engage with fans directly. Collectively, these digital alternatives complement streaming by diversifying the revenue portfolio of the industry, showcasing the multifaceted nature of music consumption in the digital age. However, the issue with this is that the complementation of streaming with TikTok has led to additional convenience for the consumer. This duo of streaming and TikTok leads to people consuming music as if it were free, as described by musician James Blake. He specifically argued that Tiktok has negatively affected songwriting by encouraging artists to write differently to appeal to the Tiktok algorithm and the attention deficit of the consumer by presenting a vast library of quick user-generated content tailor-made for the consumer.

@@

The analysis within this paper is not without limitations that should be acknowledged. Firstly, the dataset's restriction to revenue figures may overlook the varied consumer behaviors and preferences that drive the market. Future research could benefit from integrating qualitative data, such as consumer surveys, to complement the quantitative revenue analysis. Additionally, the rapid pace of technological change means that the dataset may quickly become outdated; continuous data collection and analysis will be important to maintain relevance. The dataset only includes information on North America. Incorporating worldwide data with @Keintz2008TheRI suggestions could lead to additional observations. Moreover, the current dataset does not capture the entire scope of the music industry and the revenue it generates from other streams of revenue such as live performances, merchandise sales, and licensing deals. Future iterations of this research should aim to include these aspects for a more complete view of the industry's economic outlook. The industry is also moving towards more immersive and interactive experiences, such as virtual concerts or Twitch streams, areas not covered by the present dataset. Including these emerging trends could provide valuable insights into future revenue potentials and shifts in consumer engagement.
